---
title: A dirty hack solution for packages using Depends
author: Karl Forner
license: GPL (>= 2)
tags: package namespace imports depends
summary: A dirty hack solution for packages using Depends
date: 2013-05-27
---

## Motivation

As stated in the the "Writing R Extensions", the Software for Data Analysis book (aka the R bible), packages should
whenever possible, to avoid name collision (masking) and ensure trustworthy computations. 
See [this discussion on Stackoverflow](http://stackoverflow.com/questions/8637993/better-explanation-of-when-to-use-imports-depends).

## The problem

So after adopting best practices, and only using **Imports** in my packages, the problem is that some packages _imported_
do not follow the those best practices and rely on **Depends** for their dependencies, and consequently will not find 
unless hacked.  

Say we are writing a package **MyPkg**, that uses some functions a CRAN package **A**, listing a CRAN package **B** in 
its Depends.

_MyPkg::f_
```
f <- function() a()
```
_A::a_
```
a <- function() b()
```

Executing `f()` will find the function a, explicitly imported from package A. But executing `a()` will die because 
function **b** can not be found.

## analysis

Because B is in A Depends, usually `b()` is found in the search path, where namespace **B** is attached. 
What can we do to work around this problem.
Sadly, there is nothing we can do at the package **MyPkg** level, because the problem sits in the **A** namespace.
We could of course lists B in the **MyPkg** Depends, but that is precisely what we want to avoid.

## the dirty hack solution

### the idea

I suggest that you first have a look at this very good blog 
[How R Searches and Finds Stuff](http://obeautifulcode.com/R/How-R-Searches-And-Finds-Stuff/) if you need some 
clarification about namespaces and imports.  


Here is a diagram of the current state of the problem:

```{r echo=FALSE}
suppressPackageStartupMessages(library(diagram))
par(mar = c(1, 1, 2, 1))

names <- c( "MyPkg:\nf()", "A:\nA()", "B\nb()", "MyPkg.Imports\nA::a()", "A.Imports", "B.Imports", "Base")
n  <- length(names)
M <- matrix(nrow = n, ncol = n, byrow = TRUE, data = 0)

M[4, 1] <- M[7, 4] <- M[5, 2] <- M[7, 5] <- M[6, 3] <- M[7, 6]  <- 1
coords <- coordinates(c(3, 3, 1))
plotmat(M, pos = coords, curve = 0, name = names, lwd = 1, box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)
curvedarrow(from = coords[1, ], to = coords[2, ], curve = +0.5, lty = 2, lcol = 2)
mtext(outer = TRUE, side = 3, line = -1.5, cex = 1.5, "Namespaces before")
```


We want the `a()` call, which is executed in namespace **A**, to find the **b** symbol in namespace **B**. 
To achieve this, we will re-route the namespace **A** (actually its Imports namespace) to the **B** namespace:
```
nsa <- getNamespace('A')
nsa_imp <- parent.env(nsa)
parent.env(nsa_imp) <- getNamespace('B')
```

Now our namespaces should look like this:  

```{r echo=FALSE}

par(mar = c(1, 1, 2, 1))

names <- c( "MyPkg:\nf()", "A:\nA()", "B\nb()", "MyPkg.Imports\nA::a()", "A.Imports", "B.Imports", "Base")
n  <- length(names)
M <- matrix(nrow = n, ncol = n, byrow = TRUE, data = 0)

M[4, 1] <- M[7, 4] <- M[5, 2] <- M[6, 3] <- M[7, 6]  <- 1
M[3, 5] <- 1
coords <- coordinates(c(3, 3, 1))
plotmat(M, pos = coords, curve = 0, name = names, lwd = 1, box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)
curvedarrow(from = coords[1, ], to = coords[2, ], curve = +0.5, lty = 2, lcol = 2)
mtext(outer = TRUE, side = 3, line = -1.5, cex = 1.5, "Namespaces before")
```

## Example

The above code is not straightforward to test because you need to write packages.  
This is some reproducible code that illustrates the hack.

```{r}
loadNamespace(multcomp)

```

## implementation

Just put the re-routing code in your MyPkg **.onLoad** function, defined usually in `R/zzz.R`:

```
.onLoad <- function(libname, pkgname) {
    nsa <- getNamespace('A')
    nsa_imp <- parent.env(nsa)
    parent.env(nsa_imp) <- getNamespace('B')    
}
```

### why dirty ?

First of all, it uses `parent.env<-`, and according to the documentation:
>The replacement function parent.env<- is extremely dangerous as it can be used to destructively change environments in ways >that violate assumptions made by the internal C code. It may be removed in the near future.

It is a pity because it can proved extremely useful.

### why a hack ?

Unfortunately it does not work in the general case. 
Suppose you need two packages A and D, that both depend on B and C, but with different rules:
```
A -> B -> C
D -> C -> B
```
In that case, we would re-route A to B, then B to C. And this is incompatible with the way D is designed, resulting in
wrong symbol picks if symbols with same name are both defined by B and C.

### so what ?

We'll see in a future post a probably much better solution.


_Karl Forner @ Quartz Bio_


