---
title: A dirty hack for importing packages that use Depends
author: Karl Forner
license: GPL (>= 2)
tags: package namespace imports depends
summary: A dirty hack for importing packages that use Depends
date: 2013-05-27
---

## Scope

This article is about R package development.

## Motivation

As stated in the the _Writing R Extensions_ manual and the _Software for Data Analysis_ book (aka the R bible), packages should
whenever possible use **Imports** instead of **Depends**, to avoid name collision (masking) and ensure trustworthy computations. 
See [this discussion on Stackoverflow](http://stackoverflow.com/questions/8637993/better-explanation-of-when-to-use-imports-depends).

## The problem

So after adopting best practices, and only using **Imports** in my packages, the problem is that some _imported_ packages 
do not follow those best practices and rely on **Depends** for their dependencies, and consequently will not find 
their dependencies unless hacked.  

Say we are writing a package **MyPkg**, that uses some functions from a CRAN package **A**, which lists a CRAN 
package **B** in its Depends.

_MyPkg::f_
```
f <- function() a()
```
_A::a_
```
a <- function() b()
```

_B::b_
b <- function() { ... }


Executing `f()` will find the function `a`, explicitly imported from package A. But executing `a()` will die because 
function `b` can not be found.

## analysis

Because B is listed in the Depends of package A, `b()` is normally found in the _search path_, where namespace **B** is attached. 
What can we do to work around this problem ?  
Sadly, there is nothing we can do at the  **MyPkg** package level, since the problem sits in the **A** namespace.
We could of course lists B in the **MyPkg** Depends, but that is precisely what we want to avoid.

## the dirty hack solution

### the idea

I suggest that you first have a look at this very good blog 
[How R Searches and Finds Stuff](http://obeautifulcode.com/R/How-R-Searches-And-Finds-Stuff/) if you need some 
clarification about namespaces and imports.  


Here is a diagram of the current state of the problem:

```{r echo=FALSE}
suppressPackageStartupMessages(library(diagram))
par(mar = c(1, 1, 2, 1))

names <- c( "MyPkg:\nf()", "A:\nA()", "B\nb()", "MyPkg.Imports\nA::a()", "A.Imports", "B.Imports", "Base")
n  <- length(names)
M <- matrix(nrow = n, ncol = n, byrow = TRUE, data = 0)

M[4, 1] <- M[7, 4] <- M[5, 2] <- M[7, 5] <- M[6, 3] <- M[7, 6]  <- ""
coords <- coordinates(c(3, 3, 1))
plotmat(M, pos = coords, curve = 0, name = names, lwd = 1, box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)
straightarrow(from = coords[4, ]+0.05, to = coords[2, ]-0.05, lty = 2, lcol = 2)
mtext(outer = TRUE, side = 3, line = -1.5, cex = 1.5, "Namespaces before")
```


We want the `a()` call, which is executed in namespace **A**, to find the **b** symbol in namespace **B**. 
To achieve this, we will re-route the namespace **A** (actually its Imports namespace) to the **B** namespace:
```
nsa <- getNamespace('A')                      # package A environment
nsa_imp <- parent.env(nsa)                    # package A Imports
parent.env(nsa_imp) <- getNamespace('B')      # re-route A Imports to package B
```

Now our namespaces should look like this:  

```{r echo=FALSE}

par(mar = c(1, 1, 2, 1))

names <- c( "MyPkg:\nf()", "A:\nA()", "B\nb()", "MyPkg.Imports\nA::a()", "A.Imports", "B.Imports", "Base")
n  <- length(names)
M <- matrix(nrow = n, ncol = n, byrow = TRUE, data = 0)

M[4, 1] <- M[7, 4] <- M[5, 2] <- M[6, 3] <- M[7, 6]  <- ""
M[3, 5] <- "New"
coords <- coordinates(c(3, 3, 1))
plotmat(M, pos = coords, curve = 0, name = names, lwd = 1, box.lwd = 2, cex.txt = 0.8, box.type = "circle", box.prop = 1.0)
straightarrow(from = coords[4, ]+0.05, to = coords[2, ]-0.05, lty = 2, lcol = 2)
mtext(outer = TRUE, side = 3, line = -1.5, cex = 1.5, "Namespaces after re-rerouting")
```

Notice that the link From A Imports to base has been re-routed towards package B (the arrow labelled "New")

## Example


The above code is not straightforward to test because you need to write a package.  

Here is some reproducible code that illustrates how the hack works, so that you can simulate its mechanism
right now by copy-pasting the following code in your R console.
This example uses package **multcomp** that _Depends_ on **mvtnorm**. 

```{r}
## load multcomp package, and its imports, but do not attach them in the search path
l <- loadNamespace('multcomp') # assignment to avoid annoying print

amod <- aov(breaks ~ wool + tension, data = warpbreaks)
wht <- multcomp::glht(amod, linfct = multcomp::mcp(tension = "Tukey"))
ci <- confint(wht) # mvtnorm::qmvt is not found, because mvtnorm is not attached to the search path

# let's load the mvtnorm namespace 
l <- loadNamespace('mvtnorm')
ci <- confint(wht) # mvtnorm::qmvt still not found, mvtnorm is loaded but not attached

# hack hack hack: re-route package multcomp to mvtnorm
ns1 <- getNamespace('multcomp')
ns1_imp <- parent.env(ns1)
parent.env(ns1_imp) <- getNamespace('mvtnorm')

ci <- confint(wht) # now should work
print(ci)

```

## implementation of the hack in MyPkg

Just put the re-routing code in your MyPkg **.onLoad** function, defined usually in `R/zzz.R`:

```
.onLoad <- function(libname, pkgname) {
    nsa <- getNamespace('A')
    nsa_imp <- parent.env(nsa)
    parent.env(nsa_imp) <- getNamespace('B')    
}
```

### why dirty ?

First of all, it uses `parent.env<-`, and according to the documentation:
>The replacement function parent.env<- is extremely dangerous as it can be used to destructively change environments in ways >that violate assumptions made by the internal C code. It may be removed in the near future.

It is a pity because it can be extremely useful.

### why a hack ?

Unfortunately it does not work in the general case. 
Suppose you need two packages A and D, that both depend on B and C, but with different rules:
```
A -> B -> C
D -> C -> B
```
In that case, we would re-route A to B, then B to C. And this is incompatible with the way D is designed, resulting in
wrong symbol picks if symbols with same name are both defined by B and C.

### so what ?

We'll see in a future post a probably much better solution.


_Karl Forner @ Quartz Bio_


